---
title: "Basecalling and Demultiplexing Summary Report"
author: "basecallQC"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## basecallQC Summary Report

The basecallQC summary report contains summary metrics on demultiplexing and basecalling for an Illumina Run. 

### Conversion Metrics

The yield of all samples before and after passing filter are presented in the table below. 

```{r LaneT,echo=F,fig.height=10,fig.width=10}
summaryConvStatsTable(BCLQCreport,output="html")
```

The bar plot shown below illustrates  the total Yield of samples before and after passing filter.
```{r Lane,echo=F,fig.height=10,fig.width=10}
passFilterBar(BCLQCreport,groupBy = "Sample",
              metricToPlot = "Yield")+theme_bw()


```

The boxplot shows the distribution of Yield of samples before and after passing filter.

```{r Sample,echo=F,fig.height=10,fig.width=10}
passFilterBoxplot(BCLQCreport,groupBy = "Sample",
              metricToPlot = "Yield")+theme_bw()
```

The below plot shows the Yield per Tile/Lane for the Illumina Run to allow for the identification of potentially problematic Tiles per lane. 

```{r Sample2,echo=F,fig.height=10,fig.width=10}
passFilterTilePlot(BCLQCreport,
              metricToPlot = "Yield")
```

### Demultiplexing Metrics


The demultiplexing count for all samples are shown in the below table. 

```{r table,echo=F,fig.height=10,fig.width=10}
summaryDemuxTable(BCLQCreport,output="html")
```

The below bar plot shows the results  per sample for demultiplexing of the Run and so should help visually identify any samples which are less represented than expected.

```{r tile,echo=F,fig.height=10,fig.width=10}
demuxBarplot(BCLQCreport,groupBy = "Sample")+theme_bw()
```

```{r fastqc,echo=F,fig.height=30,fig.width=15}
if(doFQQC==TRUE){
  makeFQTable(BCLQCreport)
}
```

```{r sessionInfo,echo=F,fig.height=30,fig.width=15}
sessionInfo()
```
